Visual diagnostics for constrained optimisation with application to guided tours

### 2020-10-09 paper/paper.Rmd
separate scripts from rmd, clean up repo

- name: Author 1
  thanks: The authors gratefully acknowledge ...
  affiliation: Department of YYY, University of XXX
- name: Author 2
  affiliation: Department of ZZZ, University of WWW
keywords:
- optimisation
- projection pursuit 
- guided tourr
- visual 
- diagnostics 
- R
abstract: Friedman & Tukey commented on their initial paper on projection pursuit in 1974 that "the  technique  used  for  maximising  the  projection index strongly influences both the statistical and the computational aspects of the procedure." While many projection pursuit indices have been proposed in the literature, few concerns the optimisation procedure. In this paper, we developed a system of diagnostics aiming to visually learn how the optimisation procedures find its way towards the optimum. This diagnostic system can be applied more generally to help practitioner to unveil the black-box in randomised iterative (optimisation) algorithms. An R package, ferrn, has been created to implement this diagnostic system.
bibliography: biblio.bib
preamble: >
  \usepackage{amssymb, amsmath, mathtools, dsfont, bbm, array, booktabs}
  \usepackage[ruled,vlined, linesnumbered]{algorithm2e}
output:
  bookdown::pdf_book:
    base_format: rticles::asa_article
    fig_height: 4
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,root.dir = here::here())

library(tidyverse)
library(ferrn)
library(patchwork)
library(ggrepel)
library(tourr)
library(gganimate)
```

```{r external, include = FALSE, cache = FALSE}
library(knitr)
files <- paste0("data/", list.files(here::here("data")))
purrr::walk(.x = files, ~load(here::here(.x), env = globalenv()))
read_chunk(here::here('scripts/vis-diag.R'))
```

@@

```{r tour-path, out.height="30%", out.width="60%", fig.align='center', fig.cap="\\label{tour-path}Each square (frame) represents the projected data with a corresponding basis. Blue frames have bases found by an optimisation algorithm iteratively whilst the white frames are constructed by geodesic interpolation between two blue frames."}
knitr::include_graphics(path = here::here("figures/tour_path_keynote/tour_path_keynote.001.png"))
```

@@

```{r toy-interp, fig.height = 3, fig.cap = "\\label{toy-interp} The resulting trace plot on the interpolated points has been ploted when using three different algorithms to optimise the index. The color represents the number of iteration. It can be observed that each algorithm differs in length in the optimisation and the curvature of the improvement for each algorithm also varies."}
bind_rows(holes_2d_better_max_tries %>% mutate(group = "Algorithm 1"), 
          interrupt_no %>% mutate(group = "Algorithm 2"), 
          holes_2d_better_random %>% mutate(group = "Algorithm 3")) %>% 
  mutate(group = fct_relevel(group, c("Algorithm 1", 
                                      "Algorithm 2", 
                                      "Algorithm 3"))) %>% 
  explore_trace_interp(group = group) + 
  theme(legend.position = "none") + 
  scale_color_botanical(palette = "fern")
```

Sometimes, we may be interested in exploring the points on the interpolation path since these are the points that will be played by the tour animation. Figure \ref{toy-interp} presents the interpolation of three different tour paths. The upper plot shows a desirable interpolation in each iteration with the index value being progressively and monotonically increasing.  While in the middle plot, the increases towards the target basis is not monotonical in the last two iterations and interpolated basis with higher index value can be found on the tour path. The lower plot is constructed using `search_better_random`, where a basis with smaller index value  has a probabilistic chance of being accepted and we can observe a much more involved pattern. In iteration three, the probabilistic acceptance produces a monotonically decreasing interpolation whilst in iteration five, six and seven, with also a probabilistic acceptance, the interpolation is now non-monotonical. When the target basis has a higher index value, iteration eight reaches this basis by first decreasing the index value. While the interpolation plot shows different possibilities of the change in index value on the tour path, the diagnosis of the validity of each pattern and the related optimisation algorithms will be postponed to Section \ref{application}.

@@

```{r toy-pca, fig.height = 3, fig.cap = "\\label{toy-pca}PCA plot of search geodesic colouring by info allows for better understanding of each stage in the geodesic search"}
# path1 <- holes_1d_geo$basis %>% 
#   flatten_dbl() %>% 
#   matrix(ncol = 5, byrow = TRUE)
# path2 <- holes_1d_better$basis %>% 
#   flatten_dbl() %>% 
#   matrix(ncol = 5, byrow = TRUE)
# set.seed(1)
# sphere <- geozoo::sphere.hollow(p = 5, n = 1000)$point
# theoretical <- matrix(c(0, 1, 0, 0, 0,
#                         0, -1, 0, 0, 0), nrow = 2, byrow = TRUE)
# 
# dt <- rbind(sphere, path1, path2, theoretical)
# pca <- dt %>% prcomp(scale. = TRUE)
# #summary(pca)
# pca_pred <- predict(pca) %>%
#   as_tibble() %>%
#   mutate(method = c(rep("sphere", nrow(sphere)),
#                     holes_1d_geo$method,
#                     holes_1d_better$method,
#                     rep("theoretical", 2))) %>%
#   group_by(method) %>%
#   mutate(id = row_number(),
#          id = ifelse(method == "sphere", 1, id)) %>%
#   ungroup()
# 
# p <- pca_pred %>%
#   ggplot(aes(x = PC1, y = PC2, color = method))+
#   geom_point() +
#   geom_point(data = pca_pred %>% filter(method == "theoretical"), 
#              size = 10) + 
#   geom_point(data = pca_pred %>% filter(method %in% c("search_better", "search_geodesic") & id == 1), size = 10) + 
#   theme(aspect.ratio = 1) + 
#   scale_color_botanical(palette = "cherry")
# 
# p

p <- bind_rows(holes_1d_geo, holes_1d_better) %>% 
  bind_theoretical(matrix(c(0, 1, 0, 0, 0), nrow = 5), 
                   index = tourr::holes(), raw_data = boa5) %>% 
  bind_theoretical(matrix(c(0, -1, 0, 0,0), nrow = 5), 
                   index = tourr::holes(), raw_data  = boa5)%>%
  explore_space_pca(col = method)  + 
  scale_color_botanical(palette = "cherry") + 
  theme(legend.position = "bottom")

p
```

@@

Figure \ref{toy-pca} shows the first two principal components of two search paths. Starting from the same position, two searching methods take different paths to get its final bases. There are two theoretical bases because they correspond to the same data projection $Y$ with a 180 degree rotation. What differentiate the two optimisers is that `search_better()` also involve random evaluation of points in the sphere during its sampling process while `search_geodesic()` doesn't and this stochastic rather than deterministic approach can be useful in complex scenario.

@@

```{r fig.height = 3, fig.cap = "\\label{toy-pca-aniamted}A selected number of frames from the animated PCA plot. With animation, it is easier to track the progrssion from the start to finish in each algorithm."}
# ani <- p + theme(legend.position = "none") +
#   transition_states(id) +
#   shadow_mark()
# 
# ani <- bind_rows(holes_1d_geo, holes_1d_better) %>% 
#   bind_theoretical(matrix(c(0, 1, 0, 0, 0), nrow = 5), 
#                    index = tourr::holes(), raw_data = data) %>% 
#   bind_theoretical(matrix(c(0, -1, 0, 0,0), nrow = 5), 
#                    index = tourr::holes(), raw_data  = data)%>%
#   explore_space_pca(col = method, animate = TRUE)  + 
#   scale_color_botanical(palette = "cherry") 
# 
# animate(ani, nframes = 350, device = "png",
#         renderer = file_renderer("figures/pca/",
#                                  prefix = "pca", overwrite = TRUE))

frames <- c("0002", "0058", "0078", "0117", "0172", "0350")
ani <- paste0(here::here("figures/"), "pca/", "pca", frames, ".png")
rl <- lapply(ani, png::readPNG)
gl <-  lapply(rl, grid::rasterGrob)
wrap_plots(gl) 
```

Animating the plots introduced above is useful, especially in the case of PCA polot since it shows the bases found by the optimiser in its natural order. In Figure \ref{toy-pca-animated}, There should actually be more writing here to and also the this examples may need to be changed.

@@

```{r toy-tour, fig.cap="\\label{toy-tour}A selected number of frames from the tour animation for viewing the 5D space of all the projection bases. The second frame on the top row view the space from a direction that is close to the one in PCA plot. The tour animation allows for a more holistic view of the full space in high dimensions from different angles."}
set.seed(1)
sphere <- geozoo::sphere.hollow(p = 5, n = 1000)$point
path1 <- holes_1d_better$basis %>% 
  flatten_dbl() %>% matrix(ncol = 5, byrow = TRUE)
path2 <- holes_1d_geo$basis %>% 
  flatten_dbl() %>% matrix(ncol = 5, byrow = TRUE)
theoretical <- matrix(c(0, 1, 0, 0, 0,
                        0, -1, 0, 0, 0), nrow = 2, byrow = TRUE)
start <- get_start(holes_1d_better) %>% pull(basis) %>% .[[1]] %>% matrix(nrow = 1)
dt <- rbind(sphere, path1, path2, theoretical, start)
colnames(dt) <- c(map_chr(1:5, ~paste0("x", .x)))
pal <- c("#D3D3D3",c("#524340",  #orchre
                     "#B4B754",  # green
                     "#F3B422" # yellow
                     ))
color <- c(rep(pal[1], nrow(sphere)),
           rep(pal[2], nrow(path1)),
           rep(pal[3], nrow(path2)),
           rep(pal[4], 2), 
           pal[2] # for start
           )
cex <- c(rep(1, nrow(sphere)),
         rep(1.5, nrow(path1)), 
         rep(1.5, nrow(path2)), 
         rep(5, 2), 
         5 # for start
         )

# 
# bind_rows(holes_1d_geo, holes_1d_better) %>% 
#   explore_space_tour(color = method)
#animate_xy(dt, col = color, cex = cex, tour_path = grand_tour()) # trial

# set.seed(123)
# render(
#   dt,
#   tour_path = grand_tour(),
#   dev = "png",
#   display = display_xy(col=color,cex = cex, axes = "bottomleft"),
#   rescale = FALSE,
#   frames = 500,
#   here::here("figures","tour", "tour%03d.png")
# )

frames <- c("001", "143", "078", "117", "172", "350")
ani <- paste0(here::here("figures/"), "tour/", "tour", frames, ".png")
rl <- lapply(ani, png::readPNG)
gl <-  lapply(rl, grid::rasterGrob)
wrap_plots(gl) 
```

@@

We use the same dataset as the toy example above to explore the search function `search_better` and we want to learn how the index value changes on the interpolation path for the `holes` index. From the left panel of Figure \ref{interruption}, we observe that when interpolating from the current basis to the target basis, the index value may not be monotone: we could reach a basis with a higher index value than the target basis on the interpolation path. In this sense, we would be better off using the basis with the highest index value on the interpolation path as the current basis for the next iteration (rather than using the target basis). Hence, an interruption is constructed to accept the interpolating bases only up to the one with the largest index value. After implementing this interruption, the search finds higher final index value with fewer steps as shown in the right panel of Figure \ref{interruption}.

```{r interruption,fig.asp=0.5, fig.cap = "\\label{interruption}A comparison of the trace plot with and without the interruption. When the interruption does not take place, the index value of the target basis can be smaller than the interpolated basis. The interruption forces the interpolation to finish at the interpolated basis with the highest index value in an iteration."}
p1_anno <- interrupt_no %>% filter(info == "interpolation") %>% mutate(id = row_number()) %>% filter(id %in% c(44, 60, 62)) %>% 
  mutate(anno = c("current basis", "interpolated basis", "target basis"))

p1 <- interrupt_no %>% mutate(id = row_number() - 1) %>% explore_trace_interp() + ggtitle("without interruption") + 
  geom_point(data = p1_anno) + 
  geom_label_repel(data = p1_anno, aes(label = anno), box.padding = 0.5) + ylim(0.8, 0.9) + xlim(0, 80) + 
  theme(legend.position = "none")

p2 <- interrupt_yes %>% explore_trace_interp() + ggtitle("with interruption") + ylim(0.8, 0.9) + xlim(0, 80) + 
  theme(legend.position = "none")

(p1 | p2) & scale_color_botanical(palette = "fern")

```

@@

```{r geo-polish, fig.height= 2.5, fig.cap = "\\label{geo-polish}Two-D projection on \`boa6\` data with holes index optimised by \`search_geodesic\`. The left panel shows the final projected data before polish and the right panel shows the one after. The separation of the clusters on the y axis becomes sharper after the polish."}
# set.seed(123456)
# render(
#   boa6,
#   tour_path = guided_tour(holes(), d = 2, search_f = search_geodesic),
#   dev = "png",
#   display = display_xy(axes = "bottomleft", verbose = TRUE),
#   rescale = FALSE,
#   frames = 100,
#   file = here::here("figures","geo_polish", "before%03d.png")
# )
# 
# last_basis <- get_best(holes_2d_geo)$basis %>% .[[1]]
# 
# set.seed(123456)
# render(
#   boa6,
#   tour_path =  guided_tour(holes(), d = 2, search_f = search_polish, alpha = 0.1),
#   dev = "png",
#   display = display_xy(axes = "bottomleft", verbose = TRUE),
#   start = last_basis,
#   rescale = FALSE,
#   frames = 100,
#   file = here::here("figures","geo_polish", "after%03d.png")
# )

# set.seed(123456)
# holes_2d_geo <- animate_xy(boa6,tour_path =
#                              guided_tour(holes(), d = 2,
#                                          search_f = search_geodesic),
#                            rescale = FALSE, verbose = TRUE)
# last_basis <- get_best(holes_2d_geo)$basis %>% .[[1]]
# 
# set.seed(123456)
# holes_2d_geo_polish <- animate_xy(boa6, tour_path =
#                                     guided_tour(holes(), d = 2,
#                                                 search_f = search_polish, alpha = 0.1),
#                                   rescale = FALSE, verbose = TRUE,
#                                   start = last_basis)
# 
# save(holes_2d_geo, file = here::here("data", "holes_2d_geo.rda"))
# save(holes_2d_geo_polish, file = here::here("data", "holes_2d_geo_polish.rda"))

before <- png::readPNG(here::here("figures","geo_polish", "before100.png"))
after <- png::readPNG(here::here("figures","geo_polish", "after004.png"))
gl <-  lapply(list(before, after), grid::rasterGrob)
wrap_plots(gl) 
```

@@

Figure \ref{geo-polish} presents the final projections found before and after applying `search_polish` using the demo code. Polish search improves the index value from `r get_best(holes_2d_geo)$index_val %>% format(digits = 4)`  to `r get_best(holes_2d_geo_polish)$index_val %>% format(digits = 4)` with reduction of weights on the non-informative variables, namely `V1`, `V8`, `V9` and `V10`. When appears in the projected data in Figure \ref{geo-polish}, polish works to sharpen the edges of each cluster.

@@

```{r kol-cdf,fig.asp=0.5,fig.cap = "\\label{kol-cdf}One-D projection on \`boa5\` data with noisy index \`norm_kol\` optimised by \`search_geodesic\` and \`search_better\`. The grey dashed line represents the index value of the theoreical best basis. \`search_geodesic\` fails to optimse the noisy index while \`search_better\` has made reasonable improvments to reach the index value close to the one of theoretical best basis."}
# set.seed(123456)
# kol_1d_geo <- 
#   animate_dist(boa5, tour_path = guided_tour(norm_kol(), d = 1,
#                                              search_f =  search_geodesic, max.tries = 100),
#                rescale = FALSE, verbose = TRUE)
# 
# set.seed(123456)
# kol_1d_better <- 
#   animate_dist(boa5, tour_path = guided_tour(norm_kol(), d = 1,
#                                              search_f =  search_better, max.tries = 100),
#                rescale = FALSE, verbose = TRUE)
# save(kol_1d_geo, file = here::here("data", "kol_1d_geo.rda"))
# save(kol_1d_better, file = here::here("data", "kol_1d_better.rda"))

index <- tourr::norm_kol()
theo_best_index_val <- index(as.matrix(boa5) %*% matrix(c(0, 1, 0, 0, 0), nrow = 5, ncol = 1))

bind_rows(kol_1d_geo, kol_1d_better) %>% 
  explore_trace_interp(col = method, group = method) +
  facet_wrap(vars(method), ncol = 2) + 
  geom_hline(yintercept = theo_best_index_val, alpha = 0.5, linetype = 2) + 
  scale_color_botanical() +
  theme(legend.position = "none", aspect.ratio = 1)

```

@@

Figure \ref{kol-cdf} compares the tracing plot of two algorithms: `search_geodesic` and `search_better`. This time, the interpolated path is no longer smooth when using either algorithm. It is also obvious that `search_geodesic` fails to optimise this rough index since there is barely improvment of the index value. On the other hand, `search_better` is doing a relatively good job on the optimisation for the 5-variable dataset `boa5`. The theoretical best basis [`r matrix(c(0, 1, 0, 0, 0))`] produces an index value of `r kol_1d_better %>% bind_theoretical(matrix = matrix(c(0, 1, 0, 0, 0), nrow = 5, ncol = 1), index = tourr::norm_kol(), raw_data = boa5) %>% tail(1) %>% pull(index_val)` and `search_better` finds the final basis [`r get_best(kol_1d_better)$basis %>% .[[1]] %>% format(digits = 4)`] with an index value of `r get_best(kol_1d_better)$index_val %>% format(digits = 4)`. A further polish step will give a marginal improvement of index value to `r get_best(kol_1d_better_polish)$index_val` with a basis of [`r get_best(kol_1d_better_polish)$basis %>% .[[1]] %>% format(digits = 4)`]. At this stage, the difference between the theoretical best and what has been found is likely due to simulation error since the best possible basis for a simulated data will be slightly off the theoretical best basis, which is derived based on the distributional assumption in Equation \ref{eq:sim-norm} to \ref{eq:sim-x7}.

```{r pca-better-default, fig.height = 6, fig.cap="\\label{pca-better-default}One-D projection of \`norm_kol\` index on \`boa6\` data optimised by \`search_better\` with 20 randomly generated seeds. Each point represents a basis in the original 6D space, reduced into 2D by PCA. The grey points are random bases generated from 6-D hollow sphere and the two yellow points repsents the local maximum corresponds to basis when \`V2\` is found and the global maximum where \`V7\` is found. The points produced by the seraching algorithm is colored by whether the global or local maximum is found with the interpolated bases highlighted as a path and the final basis as an amplified point."}
set.seed(123)
seed <- sample.int(10000, 20)
# kol_1d_2var_better_sim <- list()
# kol_1d_2var_better_sim_polish <- list()
# for (i in 1: length(seed)){
#   set.seed(seed[i])
#   kol_1d_2var_better_sim[[i]] <-
#     animate_dist(boa6,
#                  tour_path = guided_tour(norm_kol(), d = 1,
#                                          search_f = search_better, max.tries = 100),
#                  rescale = FALSE, verbose = TRUE)
# 
# 
#   last_basis <- get_best(kol_1d_2var_better_sim[[i]])$basis %>% .[[1]]
# 
#   set.seed(1842)
#   kol_1d_2var_better_sim_polish[[i]] <-
#     animate_dist(boa6,
#                  tour_path = guided_tour(norm_kol(), d = 1,
#                                          search_f = search_polish, max.tries = 100),
#                  rescale = FALSE, verbose = TRUE,
#                  start = last_basis)
# }

# save(kol_1d_2var_better_sim, file = here::here("data", "kol_1d_2var_better_sim.rda"))
# save(kol_1d_2var_better_sim_polish, file = here::here("data", "kol_1d_2var_better_sim_polish.rda"))

polish_list <- purrr::map2(kol_1d_2var_better_sim_polish,seed,  ~.x %>% dplyr::mutate(seed = .y)) 
better_list <- purrr::map2(kol_1d_2var_better_sim,seed,  ~.x %>% dplyr::mutate(seed = .y)) 
polish_tidy <- do.call(dplyr::bind_rows, polish_list)
better_tidy <- do.call(dplyr::bind_rows, better_list)
all_tidy_better <- bind_rows(better_tidy, polish_tidy)

best <- get_best(all_tidy_better, group = seed) %>% arrange(index_val)

all_summary <- best$basis %>% 
  purrr::flatten_dbl() %>% matrix(ncol = 6, byrow = TRUE) %>% 
  as_tibble() %>% 
  rename("V7" = "V3", "V8" = "V4", "V9" = "V5", "V10" = "V6") %>% 
  mutate(IndexVal = best$index_val, seed = as.factor(best$seed)) %>% 
  pivot_longer(-c(seed, IndexVal), names_to = "VarFound", values_to = "value") %>% 
  group_by(seed, IndexVal) %>% 
  filter(abs(value) == max(abs(value))) %>% 
  arrange(IndexVal) %>% ungroup()

pca <- all_tidy_better %>% 
  mutate(seed = as.factor(seed)) %>% 
  left_join(all_summary %>% select(seed, VarFound, IndexVal), by = "seed") %>%
  mutate(VarFound = fct_relevel(VarFound, c("V7", "V2"))) %>% 
  bind_theoretical(matrix = matrix(c(0, 1, 0, 0, 0, 0), nrow = 6),
                                        index = tourr::norm_kol(),
                                        raw_data = boa6) %>% 
   bind_theoretical(matrix = matrix(c(0, 0, 1, 0, 0, 0), nrow = 6),
                                        index = tourr::norm_kol(),
                                        raw_data = boa6) %>%

  compute_pca()

random <- pca$aug %>% filter(method == "randomly_generated") %>% select(-seed)
theoretical <- pca$aug %>% filter(info == "theoretical") %>% select(-seed)

pca$aug %>% filter(method != "randomly_generated"& info != "theoretical") %>% 
  group_by(seed) %>% arrange(IndexVal) %>% 
  ggplot(aes(x = PC1, y = PC2, col = VarFound)) + 
  geom_point(data = random, col = "grey", size = 0.3) +
  geom_point(aes(col = VarFound), alpha = 0.1) + 
  geom_point(data = theoretical, size = 5, col = botanical_palettes$cherry[3]) + 
  geom_path(data = pca$aug %>% filter(info == "interpolation"), size = 2) + 
  geom_point(data = pca$aug %>% get_best(group = seed), size = 5) +
  geom_point(data = pca$aug %>% get_start() %>% filter(method == "search_better_random"), size = 3) +
  facet_wrap(vars(fct_reorder(seed, IndexVal))) +
  theme_void() + 
  theme(aspect.ratio = 1, legend.position = "bottom") + 
  scale_color_botanical(palette = "fern")  
```

```{r pca-better-random-default, fig.height = 6, fig.cap = "\\label{pca-better-random-default}One-D projection of \`norm_kol\` index on \`boa6\` data optimised by \`search_better_random\` with the same seeds. \`search_better_random\` has a probablistic acceptance implementation that would also a basis with lower index value. This design allows the optimiser to jump out of the local maximum and hence more incidents find the true global maximum. "}
set.seed(123)
seed <- sample.int(10000, 20)
# kol_1d_2var_better_random_sim2 <- list()
# kol_1d_2var_better_random_sim_polish2 <- list()
# for (i in 1: length(seed)){
#   cat("i = ", i, "seed = ", seed[i], "\n")
#   set.seed(seed[i])
#   kol_1d_2var_better_random_sim2[[i]] <-
#     animate_dist(boa6,
#                  tour_path = guided_tour(norm_kol(), d = 1,
#                                          search_f = search_better_random, max.tries = 200),
#                  rescale = FALSE, verbose = TRUE)
# 
# 
#   last_basis <- get_best(kol_1d_2var_better_random_sim2[[i]])$basis %>% .[[1]]
# 
#   set.seed(seed[i])
#   kol_1d_2var_better_random_sim_polish2[[i]] <-
#     animate_dist(boa6,
#                  tour_path = guided_tour(norm_kol(), d = 1,
#                                          search_f = search_polish, max.tries = 200),
#                  rescale = FALSE, verbose = TRUE,
#                  start = last_basis)
# 
# }
# save(kol_1d_2var_better_random_sim2, file = here::here("data", "kol_1d_2var_better_sim2.rda"))
# save(kol_1d_2var_better_random_sim_polish2, file = here::here("data", "kol_1d_2var_better_sim_polish2.rda"))

polish_list <- purrr::map2(kol_1d_2var_better_random_sim_polish2,seed,  ~.x %>% dplyr::mutate(seed = .y)) 
better_list <- purrr::map2(kol_1d_2var_better_random_sim2, seed,  ~.x %>% dplyr ::mutate(seed = .y)) 
polish_tidy <- do.call(dplyr::bind_rows, polish_list)
better_tidy <- do.call(dplyr::bind_rows, better_list)
all_tidy <- bind_rows(better_tidy, polish_tidy)

best <- get_best(all_tidy, group = seed) %>% arrange(index_val)

all_summary <- best$basis %>% 
  purrr::flatten_dbl() %>% matrix(ncol = 6, byrow = TRUE) %>% 
  as_tibble() %>% 
  rename("V7" = "V3", "V8" = "V4", "V9" = "V5", "V10" = "V6") %>% 
  mutate(IndexVal = best$index_val, seed = as.factor(best$seed)) %>% 
  pivot_longer(-c(seed, IndexVal), names_to = "VarFound", values_to = "value") %>% 
  group_by(seed, IndexVal) %>% 
  filter(abs(value) == max(abs(value))) %>% 
  arrange(IndexVal) %>% ungroup()

pca <- all_tidy %>% 
  mutate(seed = as.factor(seed)) %>% 
  left_join(all_summary %>% select(seed, VarFound, IndexVal), by = "seed") %>%
  mutate(VarFound = fct_relevel(VarFound, c("V7", "V2"))) %>% 
  bind_theoretical(matrix = matrix(c(0, 1, 0, 0, 0, 0), nrow = 6),
                   index = tourr::norm_kol(),
                   raw_data = boa6) %>% 
  bind_theoretical(matrix = matrix(c(0, 0, 1, 0, 0, 0), nrow = 6),
                   index = tourr::norm_kol(),
                   raw_data = boa6) %>%

  compute_pca()

random <- pca$aug %>% filter(method == "randomly_generated") %>% select(-seed)
theoretical <- pca$aug %>% filter(info == "theoretical") %>% select(-seed)

>>>>

- name: Author 1
  thanks: The authors gratefully acknowledge ...
  affiliation: Department of YYY, University of XXX
- name: Author 2
  affiliation: Department of ZZZ, University of WWW
keywords:
- optimisation
- projection pursuit 
- guided tourr
- visual 
- diagnostics 
- R
abstract: Friedman & Tukey commented on their initial paper on projection pursuit in 1974 that "the  technique  used  for  maximising  the  projection index strongly influences both the statistical and the computational aspects of the procedure." While many projection pursuit indices have been proposed in the literature, few concerns the optimisation procedure. In this paper, we developed a system of diagnostics aiming to visually learn how the optimisation procedures find its way towards the optimum. This diagnostic system can be applied more generally to help practitioner to unveil the black-box in randomised iterative (optimisation) algorithms. An R package, ferrn, has been created to implement this diagnostic system.
bibliography: biblio.bib
preamble: >
  \usepackage{amssymb, amsmath, mathtools, dsfont, bbm, array, booktabs}
  \usepackage[ruled,vlined, linesnumbered]{algorithm2e}
output:
  bookdown::pdf_book:
    base_format: rticles::asa_article
    fig_height: 3
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      root.dir = here::here(), fig.path = here::here("figs/"), 
                      fig.align = 'center')
```

```{r external, include = FALSE, cache = FALSE}
library(knitr)
read_chunk(here::here('scripts/vis-diag.R'))
read_chunk(here::here('scripts/examples.R'))
read_chunk(here::here('scripts/kol-better-default.R'))
read_chunk(here::here('scripts/kol-better-random-default.R'))
read_chunk(here::here('scripts/kol-better-random-tuned.R'))
read_chunk(here::here('scripts/implementation.R'))
```

```{r load-pkg}
```

@@

```{r tour-path, out.height="30%", out.width="60%", fig.cap="(ref:tour-path)"}
include_graphics(here::here("img/tour_path.png"))
```

(ref:tour-path) Each square (frame) represents the projected data with a corresponding basis. Blue frames have bases found by an optimisation algorithm iteratively whilst the white frames are constructed by geodesic interpolation between two blue frames.

@@

```{r toy-interp,fig.cap = "(ref:toy-interp)"}
```

(ref:toy-interp) The resulting trace plot on the interpolated points has been ploted when using three different algorithms to optimise the index. The color represents the number of iteration. It can be observed that each algorithm differs in length in the optimisation and the curvature of the improvement for each algorithm also varies.

Sometimes, we may be interested in exploring the points on the interpolation path since these are the points that will be played by the tour animation. Figure \@ref(fig:toy-interp) presents the interpolation of three different tour paths. The upper plot shows a desirable interpolation in each iteration with the index value being progressively and monotonically increasing.  While in the middle plot, the increases towards the target basis is not monotonical in the last two iterations and interpolated basis with higher index value can be found on the tour path. The lower plot is constructed using `search_better_random`, where a basis with smaller index value  has a probabilistic chance of being accepted and we can observe a much more involved pattern. In iteration three, the probabilistic acceptance produces a monotonically decreasing interpolation whilst in iteration five, six and seven, with also a probabilistic acceptance, the interpolation is now non-monotonical. When the target basis has a higher index value, iteration eight reaches this basis by first decreasing the index value. While the interpolation plot shows different possibilities of the change in index value on the tour path, the diagnosis of the validity of each pattern and the related optimisation algorithms will be postponed to Section \ref{application}.

@@

```{r toy-pca, fig.cap = "(ref:toy-pca)"}
```

(ref:toy-pca) PCA plot of search geodesic colouring by info allows for better understanding of each stage in the geodesic search

@@

Figure \@ref(fig:toy-pca) shows the first two principal components of two search paths. Starting from the same position, two searching methods take different paths to get its final bases. There are two theoretical bases because they correspond to the same data projection $Y$ with a 180 degree rotation. What differentiate the two optimisers is that `search_better()` also involve random evaluation of points in the sphere during its sampling process while `search_geodesic()` doesn't and this stochastic rather than deterministic approach can be useful in complex scenario.

@@

```{r toy-pca-aniamted, fig.cap = "(ref:toy-pca-aniamted)"}
```

(ref:toy-pca-aniamted) A selected number of frames from the animated PCA plot. With animation, it is easier to track the progrssion from the start to finish in each algorithm.

Animating the plots introduced above is useful, especially in the case of PCA polot since it shows the bases found by the optimiser in its natural order. In Figure \@ref(fig:toy-pca-animated), There should actually be more writing here to and also the this examples may need to be changed.

@@

```{r toy-tour, fig.cap="(ref:toy-tour)"}
```

(ref:toy-tour) A selected number of frames from the tour animation for viewing the 5D space of all the projection bases. The second frame on the top row view the space from a direction that is close to the one in PCA plot. The tour animation allows for a more holistic view of the full space in high dimensions from different angles.

@@

We use the same dataset as the toy example above to explore the search function `search_better` and we want to learn how the index value changes on the interpolation path for the `holes` index. From the left panel of Figure \@ref(fig:interruption), we observe that when interpolating from the current basis to the target basis, the index value may not be monotone: we could reach a basis with a higher index value than the target basis on the interpolation path. In this sense, we would be better off using the basis with the highest index value on the interpolation path as the current basis for the next iteration (rather than using the target basis). Hence, an interruption is constructed to accept the interpolating bases only up to the one with the largest index value. After implementing this interruption, the search finds higher final index value with fewer steps as shown in the right panel of Figure \@ref(fig:interruption).

```{r interruption, fig.asp=0.5, fig.cap = "(ref:interruption)"}
```

(ref:interruption) A comparison of the trace plot with and without the interruption. When the interruption does not take place, the index value of the target basis can be smaller than the interpolated basis. The interruption forces the interpolation to finish at the interpolated basis with the highest index value in an iteration.

@@

```{r polish, fig.height= 2.5, fig.cap = "(ref:polish)"}
```

(ref:polish) Two-D projection on `boa6` data with holes index optimised by `search_geodesic`. The left panel shows the final projected data before polish and the right panel shows the one after. The separation of the clusters on the y axis becomes sharper after the polish.

@@

Figure \@ref(fig:polish) presents the final projections found before and after applying `search_polish` using the demo code. Polish search improves the index value from `r get_best(holes_2d_geo)$index_val %>% format(digits = 4)`  to `r get_best(holes_2d_geo_polish)$index_val %>% format(digits = 4)` with reduction of weights on the non-informative variables, namely `V1`, `V8`, `V9` and `V10`. When appears in the projected data in Figure \@ref(fig:polish), polish works to sharpen the edges of each cluster.

@@

```{r noisy-better-geo,fig.asp=0.5,fig.cap = "(ref:noisy-better-geo)"}
```

(ref:noisy-better-geo) One-D projection on `boa5` data with noisy index `norm_kol` optimised by `search_geodesic` and `search_better`. The grey dashed line represents the index value of the theoreical best basis. `search_geodesic` fails to optimse the noisy index while `search_better` has made reasonable improvments to reach the index value close to the one of theoretical best basis.

@@

Figure \@ref(fig:noisy-better-geo) compares the tracing plot of two algorithms: `search_geodesic` and `search_better`. This time, the interpolated path is no longer smooth when using either algorithm. It is also obvious that `search_geodesic` fails to optimise this rough index since there is barely improvment of the index value. On the other hand, `search_better` is doing a relatively good job on the optimisation for the 5-variable dataset `boa5`. The theoretical best basis [`r matrix(c(0, 1, 0, 0, 0))`] produces an index value of `r kol_1d_better %>% bind_theoretical(matrix = matrix(c(0, 1, 0, 0, 0), nrow = 5, ncol = 1), index = tourr::norm_kol(), raw_data = boa5) %>% tail(1) %>% pull(index_val)` and `search_better` finds the final basis [`r get_best(kol_1d_better)$basis %>% .[[1]] %>% format(digits = 4)`] with an index value of `r get_best(kol_1d_better)$index_val %>% format(digits = 4)`. A further polish step will give a marginal improvement of index value to `r get_best(kol_1d_better_polish)$index_val` with a basis of [`r get_best(kol_1d_better_polish)$basis %>% .[[1]] %>% format(digits = 4)`]. At this stage, the difference between the theoretical best and what has been found is likely due to simulation error since the best possible basis for a simulated data will be slightly off the theoretical best basis, which is derived based on the distributional assumption in Equation \ref{eq:sim-norm} to \ref{eq:sim-x7}.

```{r kol-better-default, fig.height = 6, fig.cap="(ref:kol-better-default)"}
```

(ref:kol-better-default) One-D projection of `norm_kol` index on `boa6` data optimised by `search_better` with 20 randomly generated seeds. Each point represents a basis in the original 6D space, reduced into 2D by PCA. The grey points are random bases generated from 6-D hollow sphere and the two yellow points repsents the local maximum corresponds to basis when `V2` is found and the global maximum where `V7` is found. The points produced by the seraching algorithm is colored by whether the global or local maximum is found with the interpolated bases highlighted as a path and the final basis as an amplified point.

```{r kol-better-random-default, fig.height = 6, fig.cap = "(ref: kol-better-random-default)"}
```

(ref: kol-better-random-default) One-D projection of `norm_kol` index on `boa6` data optimised by `search_better_random` with the same seeds. `search_better_random` has a probablistic acceptance implementation that would also a basis with lower index value. This design allows the optimiser to jump out of the local maximum and hence more incidents find the true global maximum.

```{r kol-better-random-tuned, fig.height= 6, fig.cap = "(ref:kol-better-random-tuned)"}
```

(ref:kol-better-random-tuned) One-D projection of `norm_kol` index on `boa6` data optimised by `search_better_random` with the same seeds. Further tuning of the parameter has been taken place to adjust the neighbourhood parameter `alpha` from 0.5 to 0.7. This time, all the 20 incidents have found the global maximum.
